### 2015-06-21 chapters/ipa_meets_unicode.tex

By Unicode Canonical Combining Classes, the diacritics on top of a character,
like <\dia{0334}>, always come before diacritics below, which in turn always
come before diacritics above. We follow this order and add the other IPA
diacritics (which are not diacritics in the Unicode sense) after these
diacritics. Within these classes of diacritics there is no canonical ordering
specified by Unicode, so we propose an explicit ordering here.

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] nasalized <\dia{0303}>
	  \item[→] centralized <\dia{0308}> \textbar{} mid-centralized <\dia{033D}>
	  \item[→] extra short <\dia{0306}>
	  \item[→] no audible release <\dia{031A}>
 \end{itemize} \end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] rhotic hook <\dia{02DE}>
	  \item[→] lateral release <\dia{02E1}> \textbar{} nasal release <\dia{207F}>
	  \item[→] labialized <\dia{02B7}>
	  \item[→] palatalized <\dia{02B2}>
	  \item[→] velarized <\dia{02E0}>
	  \item[→] pharyngealized <\dia{02E4}>
	  \item[→] aspirated <\dia{02B0}> \textbar{} ejective <\dia{02BC}>
	  \item[→] long <\dia{02D0}> \textbar{} half-long <\dia{02D1}>
%	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
	\end{itemize}
\end{itemize}

\noindent Finally, Tone markers are put at the end of any such sequence:

\begin{itemize}
  \item[] \textsc{Tone Marks:}
  \begin{itemize}
    \item[→] tone letters <˥ ˦ ˧ ˨ ˩> \textbar{} tone movement <↑ ↓ ↗ ↘>
  \end{itemize}
\end{itemize}

@@

At the end of this chapter we have added a few longish tables summarizing all
157 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 43 diacritics and 7 remaining symbols). Each of
these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol.

@@

  \item \textsc{IPA letters} \newline
        The 107 different IPA letters as allowed in strict-IPA encoding are
        listed in Table~\ref{tab:ipa_letters} starting on
        page~\pageref{tab:ipa_letters}.
  \item \textsc{IPA diacritics} \newline The 43 different IPA diacritics and
        tone markers (both Unicode Modifier Letters and Combining Diacritical
        Marks) as allowed in strict-IPA encoding are listed in
        Table~\ref{tab:ipa_diacritics} starting on
        page~\pageref{tab:ipa_diacritics}.
  \item \textsc{IPA remainders} \newline
        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
        page~\pageref{tab:ipa_boundaries}. 
  \item \textsc{IPA additions} \newline
        The 10 proposed additions to the IPA are listed in
        Table~\ref{tab:ipa_additions} on page~\pageref{tab:ipa_additions}.

>>>>

By Unicode Canonical Combining Classes, the diacritics on top of a character,
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Within all these classes of diacritics there is no canonical
ordering specified by Unicode, so we propose an explicit ordering here.

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] nasalized <\dia{0303}>
	  \item[→] centralized <\dia{0308}> \textbar{} mid-centralized <\dia{033D}>
	  \item[→] extra short <\dia{0306}>
	  \item[→] no audible release <\dia{031A}\ >
 \end{itemize} \end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] rhotic hook <\dia{02DE}>
	  \item[→] lateral release <\dia{02E1}> \textbar{} nasal release <\dia{207F}>
	  \item[→] labialized <\dia{02B7}>
	  \item[→] palatalized <\dia{02B2}>
	  \item[→] velarized <\dia{02E0}>
	  \item[→] pharyngealized <\dia{02E4}>
	  \item[→] aspirated <\dia{02B0}> \textbar{} ejective <\dia{02BC}>
	  \item[→] long <\dia{02D0}> \textbar{} half-long <\dia{02D1}>
	\end{itemize}
\end{itemize}

\noindent Finally, Tone markers are put at the end of any such sequence, and the 
tie bar follows at the very end:

\begin{itemize}
  \item[] \textsc{Tone Marks:}
  \begin{itemize}
    \item[→] tone letters <˥ ˦ ˧ ˨ ˩> \textbar{} tone movement <↑ ↓ ↗ ↘>
    \item[→] tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
  \end{itemize}
\end{itemize}

@@

At the end of this chapter we have added a few longish tables summarizing all
159 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 45 diacritics and 7 remaining symbols). Each of
these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol.

@@

  \item \textsc{IPA letters} \newline
        The 107 different IPA letters as allowed in strict-IPA encoding are
        listed in Table~\ref{tab:ipa_letters} starting on
        page~\pageref{tab:ipa_letters}.
  \item \textsc{IPA diacritics} \newline The 45 different IPA diacritics and
        tone markers (both Unicode Modifier Letters and Combining Diacritical
        Marks) as allowed in strict-IPA encoding are listed in
        Table~\ref{tab:ipa_diacritics} starting on
        page~\pageref{tab:ipa_diacritics}.
  \item \textsc{IPA remainders} \newline
        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
        page~\pageref{tab:ipa_boundaries}. 
  \item \textsc{IPA additions} \newline
        The 10 proposed additions to the IPA are listed in
        Table~\ref{tab:ipa_additions} on page~\pageref{tab:ipa_additions}.

### 2015-06-21 chapters/orthography_profiles.tex

>>>>

\section{Examples}

[Here should a few abstract short simple examples be added]

\

Note that to deal with ambiguous parsing cases, we can use the Unicode approach
using the zero width joiner. This is actually a non-joiner (the name is
confusing): the idea is to add this character into the text to identify cases in
which a sequence of characters is not supposed to be a complex grapheme (even
though the sequence is in the orthography profile)

### 2015-06-21 chapters/pitfalls.tex

Inspecting the Unicode Standard, there appear to be special Unicode characters
to ``glue'' together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called ``collation''
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures. To solve the issue
of tailored grapheme clusters, Unicode offers some assistance in the form of the
Unicode Locale Descriptions. However, in the practice of linguistic research,
this is not a real solution. For that reason we propose to use orthography
profiles (see Chapter~\ref{orthography-profiles}).

>>>>

Inspecting the Unicode Standard, there appear to be special Unicode characters
to ``glue'' together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called ``collation''
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures.

To solve the issue of tailored grapheme clusters, Unicode offers some assistance
in the form of the Unicode Locale Descriptions. However, in the practice of
linguistic research, this is not a real solution. For that reason we propose to
use orthography profiles (see Chapter~\ref{orthography-profiles}). Basically,
both orthography profiles and locale descriptions offer a way to specify
tailored grapheme clusters. For example, for English one could specify that <sh>
is such a cluster. Consequently, this sequence of characters is then always
interpreted as a complex grapheme. For cases in which this is not the right
decision, like in the English word \textit{mishap}, the \textsc{zero width
joiner} at \uni{200D} has to be entered between <s> and <h>.

### 2015-06-21 tables/unicode_boundaries.tex

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
\charis{‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}

>>>>

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.4cm} L{4cm} }
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
 & \unif{0020} & \textsc{space} & word break \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
\charis{‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}
